---
title: "Week 2 quiz"
output: html_document
date: "2024-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 2

```{r }
library(GSE5859Subset)
data(GSE5859Subset)
```

Compute the SVD of geneExpression. Save it as the variable s.
A: What is the first entry of s$d?
```{r }
s <- svd(geneExpression)
s1 <- s$d[1]
s1
```
B: The proportion of variabiilty in the data explained by the xth column of U is equal to s$d[x]^2 divided by the sum of all s$d values. What proportion of variability is explained by the first column of U?
```{r }
U <- s$d[1]^2/ sum(s$d^2)
U
```
C: Compute the mean of each row of geneExpression as a vector m. What is the correlation between m and the first column of s$u?
```{r }
m <- rowMeans(geneExpression)
cor(m,s$u[,1])
```

## Question 4

Define y as geneExpression - rowMeans(geneExpression), then compute the SVD of y and save the result as s.

A: What is the first entry of s$d?
```{r }
y <- geneExpression - rowMeans(geneExpression)
s <- svd(y)
s$d[1]
```
B: What proportion of variability is explained by the first column of U ?
```{r }
U <- s$d[1]^2/ sum(s$d^2)
U
```
C: Calculate the proportion of variability explained by each column of U. How many individual columns explain more than 5% of the variability?
```{r }
variability_proportion <- (s$d^2) / sum(s$d^2)
sum(variability_proportion > 0.05)
```
D: What percent of variability is explained by the first 10 rows of U? Please enter in proportion terms; if your answer is 25%, then enter .25 here.
```{r }
per_10 <- sum(s$d[1:10]^2) / sum(s$d^2)
per_10
```

## Question 5

```{r }
s <- svd(geneExpression)
y2 <- s$u %*% diag(s$d) %*% t(s$v)
residuals <- geneExpression - y2
max(abs(residuals))
```

## Question 6

Let z = s$d * t(s$v). Compare the distance between columns 1 and 2 in geneExpression (the original matrix), y (the de-trended matrix with row means subtracted), and z.

A: What is the distance between columns 1 and 2 in geneExpression?
```{r }
sqrt(crossprod(geneExpression[,1]-geneExpression[,2]))
```

B: What is the distance between columns 1 and 2 in y?
```{r }
sqrt(crossprod(y[,1]-y[,2]))
```
C: What is the distance between columns 1 and 2 in z?
```{r }
z = s$d * t(s$v)
sqrt(crossprod(z[,1]-z[,2]))
```
D: What is the distance between columns 1 and 2 in z using only the first 10 rows as an approximation?
```{r }
sqrt(crossprod(z[1:10,1]-z[1:10,2]))
```

## Question 7
Perform MDS on the original geneExpression data:

```{r }
d = dist(t(geneExpression))
mds = cmdscale(d)
fdate = factor(sampleInfo$date)
plot(mds[,1], mds[,2], col = fdate)
legend("topright",levels(fdate),col=seq_along(fdate),pch=1)
```
