---
title: "Week 2"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SVD

```{r }
library(rafalib)
library(MASS)
n <- 100
set.seed(1)
y <- t(mvrnorm(n,c(0,0),
               matrix(c(1,0.95,0.95,1),
                      2.2)))
mypar()
LIM <- c(-3.5,3.5)
plot(y[1,],y[2,], xlim=LIM, ylim=LIM)
```
## now apply singular value decomposition (SVD)

```{r }
s <- svd(y)
PC1 <- s$d[1]*s$v[,1]
PC2 <- s$d[2]*s$v[,2]
plot(PC1,PC2,xlim=LIM,ylim=LIM)
```

```{r }
library(tissuesGeneExpression)
data(tissuesGeneExpression)
dim(e)
```

```{r }
set.seed(1)
ind <- sample(nrow(e),500) #only we will take 500 rows sample
Y <- t(apply(e[ind,],1,scale)) #standardize the data
s <- svd(Y)
U <- s$u
V <- s$v
D <- diag(s$d)
Yhat <- U %*% D %*% t(V)
resid <- Y - Yhat
max(abs(resid))
```

## if we have anything very close to zero, we will remove it
```{r }
plot(s$d)
```
we see that there are some values close to zero 
```{r }
k <- ncol(U) - 4 #removing last four columns 
Yhat <- U[,1:k] %*% D[1:k,1:k] %*% t(V[,1:k])
resid <- Y - Yhat
max(abs(resid))
```
so after removing last 4 columns , we keep it the same so it's a good reduction 

```{r }
plot (s$d^2 / sum(s$d^2)* 100)
```
```{r }
k <- ncol(U) - 95  #reduce dimension by half
Yhat <- U[,1:k] %*% D[1:k,1:k] %*% t(V[,1:k])
resid <- Y - Yhat
boxplot(resid,ylim=LIM)
```
we are close to y, even when we reduce the dimensions by a half

we will take variance of residuals and divide it by variance of original y
```{r }
var(as.vector(resid))/var(as.vector(Y))
```
we only lost 4%

another way to do it
```{r }
1-sum(s$d[1:k]^2)/sum(s$d^2)
```

we will create a simulation where we have highly correlated data
```{r }
m <- 100
n <- 2
x <- rnorm(m)
e <- rnorm(n*m,0,0.01)
y <- cbind(x,x)+e
cor(y)
```

```{r }
svd(y)$d
```
we can see from svd that there are a lot of variability in the data

## Projections Exercises
```{r }
library(Biobase)
library(GSE5859Subset)
data(GSE5859Subset)
```
## SVD Exercises

```{r }
library(tissuesGeneExpression)
data(tissuesGeneExpression)
```

## SVD Exercises #1

Compute the SVD of e and the mean of each row:
```{r }
s = svd(e)
m = rowMeans(e)
```

What is the correlation between the first column of U and m?
```{r }
cor(s$u[,1],m)
```

## SVD Exercises #2

In the above question, we saw how the first column relates to the mean of the rows of e. Note that if we change these means, the distances between columns do not change. Here is some R code showing how changing the means does not change the distances
```{r }
newmeans = rnorm(nrow(e)) #random values we will add to create
newe = e+newmeans #change the mean
sqrt(crossprod(e[,3]-e[,45]))
sqrt(crossprod(newe[,3]-newe[,45]))
```
So we might as well make the mean of each row 0 since it does not help us approximate the column distances. We will define y as the detrended e and recompute the SVD:
```{r }
y = e - rowMeans(e)
s = svd(y)
```

We showed that UDV^T  is equal to y up to numerical error:
```{r }
resid = y - s$u %*% diag(s$d) %*% t(s$v)
max(abs(resid))
```

The above can be made more efficient in two ways. First, using the crossprod() and second not creating a diagonal matrix. Note that in R we can multiply a matrix x by vector a. The result is a matrix with row i equal to x[i,]*a[i]. Here is an example to illustrate this.
```{r }
x = matrix(rep(c(1,2),each=5),5,2)
x
x*c(1:5)
```
Note that the above code is actually equivalent to:
```{r }
sweep(x,1,1:5,"*")
```
This means that we don't have to convert s$d into a matrix to obtain DV^T

Which of the following gives us the same as diag(s$d)%*%t(s$v)?
A : s$d * t(s$v)

## SVD Exercises #3
If we define vd = t(s$d * t(s$v)), then which of the following is not the same as UDV^T :
A: s$u %*% s$d * t(s$v)

## SVD Exercises #4
Let z = s$d * t(s$v). We showed a derivation demonstrating that because U is orthogonal, the distance between e[,3] and e[,45] is the same as the distance between y[,3] and y[,45], which is the same as z[,3] and z[,45]:
```{r }
z = s$d * t(s$v)
sqrt(crossprod(e[,3]-e[,45]))
sqrt(crossprod(y[,3]-y[,45]))
sqrt(crossprod(z[,3]-z[,45]))
```
Note that the columns z have 189 entries, compared to 22,215 for e.

What is the difference (in absolute value) between the actual distance sqrt(crossprod(e[,3]-e[,45])) and the approximation using only two dimensions of z?
```{r }
realdistance = sqrt(crossprod(e[,3]-e[,45]))
approxdistance = sqrt(crossprod(z[1:2,3]-z[1:2,45]))
abs(realdistance - approxdistance)
```

## SVD Exercises #5

What is the minimum number of dimensions we need to use for the approximation in SVD Exercises #4 to be within 10% or less?
```{r }
ks = 1:189
realdistance = sqrt(crossprod(e[,3]-e[,45]))
approxdistances = sapply(ks,function(k){
    sqrt(crossprod(z[1:k,3,drop=FALSE]-z[1:k,45,drop=FALSE] )) 
  })
percentdiff = 100*abs(approxdistances - realdistance)/realdistance
plot(ks,percentdiff) ##take a look
min(ks[which(percentdiff < 10)])
```
## SVD Exercises #6

Compute distances between sample 3 and all other samples:
```{r }
distances = sqrt(apply(e[,-3]-e[,3],2,crossprod))
```

Recompute this distance using the 2 dimensional approximation.

What is the Spearman correlation between this approximate distance and the actual distance?

```{r }
approxdistances = sqrt(apply(z[1:2,-3]-z[1:2,3],2,crossprod))
plot(distances,approxdistances) ##take a look
```

```{r }
cor(distances,approxdistances,method="spearman")
```

## Multi-Dimensional Scaling Plots (MDS)

we will look at 3 tissues kidney, colon, and liver 
```{r }
library(rafalib)
library(tissuesGeneExpression)
data(tissuesGeneExpression)
colind <- tissue%in%c("kidney","colon","liver")
mat <- e[,colind]
group <- factor(tissue[colind])
dim(mat)
```

we can't make picture of two points that i have 22,000
we will Compute SVD
```{r }
#install.packages("ftissue")
s <- svd(mat - rowMeans(mat))
z <- diag(s$d[1:2]) %*% t(s$v[,1:2]) # z matrix
z <- t(z)
#plot(z)

library(rafalib)
mypar(1,1)
#plot(z)

plot(z[,1],z[,2],bg=as.numeric(ftissue),pch=21,xlab="First dimension",ylab="Second dimension")
legend("bottomright", levels(ftissue),col=seq(along=levels(ftissue)),pch=15)
```
## MDS Exercises #1

```{r }
library(tissuesGeneExpression)
data(tissuesGeneExpression)
```

In these exercise we will demonstrate the relantionship between the SVD and the output of cmdscale(), the function in R that performs MDS.

Using the z we computed in SVD Exercises #4:
```{r }
y = e - rowMeans(e)
s = svd(y)
z = s$d * t(s$v)
```

we can make an MDS plot:
```{r }
library(rafalib)
ftissue = factor(tissue)
mypar(1,1)
plot(z[1,],z[2,],col=as.numeric(ftissue))
legend("topleft",levels(ftissue),col=seq_along(ftissue),pch=1)
```
Now run the function cmdscale() on the original data:
```{r }
d = dist(t(e))
mds = cmdscale(d)
```

What is the correlation between the first row of z and the first column in mds?
```{r }
cor(z[1,],mds[,1])
```

## MDS Exercises #2

What is the correlation between the second row of z and the second column of mds?
```{r }
cor(z[2,],mds[,2])

```

## MDS Exercises #3

Note that the MDS plot is not the same:
```{r }
library(rafalib)
ftissue = factor(tissue)
mypar(1,2)
plot(z[1,],z[2,],col=as.numeric(ftissue))
legend("topleft",levels(ftissue),col=seq_along(ftissue),pch=1)
plot(mds[,1],mds[,2],col=as.numeric(ftissue))
```
Given the answer to MDS Exercises #1 and #2, what do we have to do to z[1,] and z[2,] to get a practically identical plot?
A: multiply z[1,] and z[2,] by -1

## MDS Exercises #4

```{r }
library(GSE5859Subset)
data(GSE5859Subset)
```

Compute the SVD and compute z:
```{r }
s = svd(geneExpression-rowMeans(geneExpression))
z = s$d * t(s$v)
```

Which dimension of z most correlates with the outcome sampleInfo$group?
```{r }
which.max(cor(sampleInfo$g,t(z)))
```

## MDS Exercises #5

Continue working with the z calculated from the GSE5859Subset data.

What is this max correlation?
```{r }
max(cor(sampleInfo$g,t(z)))
```

## MDS Exercises #6
Continue working with the z calculated from the GSE5859Subset data.

Which dimension of z has the second highest correlation with the outcome sampleInfo$group?
```{r }
which.max(cor(sampleInfo$g,t(z))[-1]) + 1 #We add 1 because we took out the first.

```

## MDS Exercises #7
Note these measurements were made during two months:
```{r }
sampleInfo$date
```
We can extract the month this way:
```{r }
month = format( sampleInfo$date, "%m")
month = factor( month)
```
Which dimension of z has the highest correlation with the outcome month?
```{r }
which.max(cor(as.numeric(month),t(z)))
```
What is this correlation?
```{r }
max(cor(as.numeric(month),t(z)))
```

## MDS Exercises #8 (ADVANCED)
Note: this is an advanced question. Please feel free to research this question online.

In MDS Exercises #7 we saw that that one of the dimensions was highly correlated to the sampleInfo$group. Now take the 5th column of U and stratify by the gene chromosome. Remove chrUn and make a boxplot of the values of U6 stratified by chromosome.

Which chromosome looks different from the rest?
Copy and paste the name as it appears in geneAnnotation.
```{r }
result = split(s$u[,6],geneAnnotation$CHR)
result = result[ which(names(result)!="chrUn") ]
boxplot(result,range=0)
```

```{r }
boxplot(result,range=0,ylim=c(-0.025,0.025))
```

```{r }
medians = sapply(result,median)
names(result)[ which.max(abs(medians)) ]
```

```{r }

```

```{r }

```

```{r }

```

```{r }

```
